A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-12-21 inst/manuscript/manuscript.tex
minor updates

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    The design-based and model-based approaches to frequentist statistical
    inference lie on fundamentally different foundations. In the
    design-based approach, inference depends on random sampling. In the
    model-based approach, inference depends on distributional assumptions.
    We compare the approaches for finite population spatial data.
  \item
    We provide relevant background for the design-based and model-based
    approaches and then study their performance using simulations and an
    analysis of real mercury concentration data. In the simulations, a
    variety of sample sizes, location layouts, dependence structures, and
    response types are considered. In the simualtions and real data
    analysis, the population mean is the parameter of interest and
    performance is measured using statistics like bias, squared error, and
    interval coverage.
  \item
    When studying the simulations and mercury concentration data, we found
    that regardless of the strength of spatial dependence in the data,
    sampling plans that incorporate spatial locations (spatially balanced
    samples) generally outperform sampling plans that ignore spatial
    locations (non-spatially balanced samples). We also found that
    model-based analyses tend to outperform design-based analyses, even
    when the data are skewed (and by consequence, the model-based
    distributional assumptions violated). The performance gap between the
    analysis approaches is small when spatially balanced samples are used
    but large when non-spatially balanced samples are used. This suggests
    that the sampling choice (spatially balanced samples versus
    non-spatially balanced samples) is most important when using a
    design-based analysis.
  \item
    There are many benefits and drawbacks practitioners must consider when
    choosing between design-based and model-based approaches for finite
    population spatial data. We provide relevant background
    contextualizing each approach and study their properties in a variety
    of scenarios, making recommendations for use based on the
    practitioner's goals.
  \end{enumerate}
  \end{abstract}

@@

There are two general approaches for using data to make frequentist
statistical inferences about a population: design-based and model-based.
When data cannot be collected for all units in a population (i.e.,
population units), data are collected on a subset of the population
units. This subset is called a sample. In the design-based approach,
inferences about the underlying population are informed via a
probabilistic process assigning some population units to be part of the
sample. Alternatively, in the model-based approach, inferences are made
from specific assumptions about the underlying process generating the
data. Each paradigm has a deep historical context (Sterba, 2009) and its
own set of benefits and drawbacks (Hansen et al., 1983).

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as data that incorporates the
specific locations of the population units into either the design or
estimation process. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Since then, there have been several general
comparisons between design-based and model-based approaches for spatial
data (Brus and De Gruijter, 1997; Brus, 2021; Ver Hoef, 2002, 2008; Wang
et al., 2012). Cooper (2006) reviews the two approaches in an ecological
context before introducing a ``model-assisted'' variance estimator that
combines aspects from each approach. In addition to Cooper (2006), there
has been substantial research and development into estimators that use
both design and model-based principles (see e.g., Sterba (2009),
Cicchitelli and Montanari (2012), Chan-Golston et al. (2020) for a
Bayesian approach).

Certainly comparisons between design-based and model-based approaches to
spatial data have been studied. But no numerical comparison has been
made between design-based approaches incorporating spatial information
and design-based approaches. In this manuscript, we compare design-based
approaches incorporating spatial information to model-based approaches
for spatial data. We focus on finite populations, but these comparisons
generalize to infinite populations as well. A finite population contains
a finite number of population units; an example is lakes (treated as a
whole with the lake centroid representing location) in the contiguous
United States. An infinite population contains an infinite number of
population units; an example is locations within a single lake.

The rest of the manuscript is organized as follows. In Section
\ref{sec:background}, we introduce and provide relevant background for
the design-based and model-based approaches to finite population spatial
data. In Section \ref{sec:mm}, we describe how we compare performance of
the approaches wiht a simulation study and an analysis of real data that
contains mercury concentration in lakes from the contiguous United
States. In Section \ref{sec:results}, we present results from the
simulation study and the analysis of mercury concentrations. And in
Section \ref{sec:discussion}, we end with a discussion and provide
directions for future research.

@@

The design-based approach assumes the population is fixed. Randomness is
incorporated via the selection of units in a sampling frame. A sampling
frame is the set of all units available to be sampled. Units from the
sampling frame are selected as part of the sample according to a
sampling design, which assigns a positive probability of inclusion
(inclusion probability) to each unit from the sampling frame. Some
examples of commonly used sampling designs include simple random
sampling, stratified random sampling, and cluster sampling. When
sampling designs incorporate spatial locations into sampling, we call
the resulting samples ``spatially balanced.'' One approach to selecting
spatially balanced samples is the Generalized Random Tessellation
Stratified (GRTS) algorithm (Stevens and Olsen, 2004), which we discuss
in more detail in Section \ref{subsec:spb_design}. When sampling designs
do not incorporate spatial locations into sampling, we call the
resulting samples ``non-spatially balanced.''

@@

The model-based approach assumes the data are a random realization of a
data-generating stochastic process. Randomness is incorporated through
distributional assumptions on this process. Strictly speaking,
randomness need not be incorporated through random sampling, though
Diggle et al. (2010) warn against preferential sampling. Preferential
sampling occurs when the process generating the data locations and the
process being modeled are not independent of one another. To guard
against preferential sampling, model-based approaches often still
implement some form of random sampling.

@@

We previously mentioned that the design-based approach can be used to
select spatially balanced samples (samples that incorporate spatial
locations of the population units and are ``well-spread'' is space).
Spatially balanced samples are useful because parameter estimates from
these samples tend to vary less than parameter estimates from samples
that are not spatially balanced (Barabesi and Franceschi, 2011;
Benedetti et al., 2017; Grafström and Lundström, 2013; Robertson et al.,
2013; Stevens and Olsen, 2004; Wang et al., 2013). The first spatially
balanced sampling algorithm seeing widespread use is the Generalized
Random Tessellation Stratified (GRTS) algorithm (Stevens and Olsen,
2004). To quantify the spatial balance of a sample, Stevens and Olsen
(2004) proposed loss metrics based on Voronoi polygons (Dirichlet
Tessellations). After the GRTS algorithm was developed, several other
spatially balanced sampling algorithms emerged, such as the Local
Pivotal Method (Grafström et al., 2012; Grafström and Matei, 2018),
Spatially Correlated Poisson Sampling (Grafström, 2012), Balanced
Acceptance Sampling (Robertson et al., 2013), Within-Sample-Distance
Sampling (Benedetti and Piersimoni, 2017), and Halton Iterative
Partitioning Sampling (Robertson et al., 2018). In this manuscript, we
select spatially balanced samples using the Generalized Random
Tessellation Stratified (GRTS) algorithm because it has several
attractive properties. More specifically, the GRTS algorithm
accommodates finite and infinite sampling frames, equal, unequal, and
proportional (to size) inclusion probabilities, legacy (historical)
sampling (Foster et al., 2017), a minimum distance between units in a
sample, and replacement units (replacement units are population units
that can be sampled when a population unit originally selected can no
longer be sampled). The GRTS algorithm selects samples by utilizing a
particular mapping between two-dimensional and one-dimensional space
that preserves proximity relationships. Via this mapping, units in
two-dimensional space are partitioned using a hierarchical address. This
hierarchical address is used to map population units to a
one-dimensional line. On the one dimensional line, each population
unit's line length equals its inclusion probability. Then, a systematic
sample of population units is selected on the line, yielding desired
sample. Stevens and Olsen (2004) provides more technical details.

After selecting a sample and collecting data, unbiased estimates of
population means and totals can be obtained using the Horvitz-Thompson
estimator (Horvitz and Thompson, 1952). If \(\tau\) is a population
total, the Horvitz-Thompson estimate of \(\tau\), denoted by
\(\hat{\tau}_{ht}\), is is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) is the value of the \(i\)th population unit in
the sample and \(\pi_i\) is the inclusion probability of the \(i\)th
population unit in the sample. An estimate of the population mean is
obtained by dividing \(\hat{\tau}_{ht}\) by \(N\), the number of
population units.

@@

Finite Population Block Kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
stochastic process. We summarize some of the basic principles of FBPK
next (for more technical details, see Ver Hoef (2008)) Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to use a sample to predict some linear function of the response
variable, \(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights (e.g, the
population mean is represented by a weights vector whose elements all
equal one). Denoting quantities that are part of the sampled population
units with a subscript \emph{s} and quantities that are part of the
unsampled population units with subscript \emph{u}, let

@@

FBPK assumes \(\bm{\delta}\) in Equation\(~\)\ref{equation:Zmarginal}
has mean-zero and a spatial correlation structure that can be modeled
using a covariance function. This covariance function is commonly
assumed to be non-negative (between zero and one), second-order
stationary (depending only on the distance between population units),
isotropic (independent of direction), and decay with distance between
population units (Cressie, 1993). Henceforth, it is implied that we have
made these same assumptions regarding \(\bm{\delta}\), though Chiles and
Delfiner (1999), pp.~80-93 discuss covariance functions that are not
second-order stationary, not isotropic, or both. A variety of flexible
covariance functions can be used to model \(\bm{\delta}\) (Cressie,
1993); one example is the exponential covariance function (for a
thorough list of spatial covariance functions, see Cressie (1993). The
\(i,j\)th element of the exponential covariance matrix,
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\), is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is the variance parameter quantifying
the variability that is dependent (coarse-scale), \(\sigma^2_{2}\) is
the variance parameter quantifying the variability that is independent
(fine-scale), \(\phi\) is the range parameter measuring the
distance-decay rate of the covariance, and \(h_{i,j}\) is the Euclidean
distance between population units \(i\) and \(j\). The proportion of
variability attributable to dependent random error is
\(\sigma^2_{1} / (\sigma^2_{1} + \sigma^2_{2})\). Similarly, the
proportion of variability attributable to independent random error is
\(\sigma^2_{2} / (\sigma^2_{1} + \sigma^2_{2})\). Finally we note that
\(\sigma^2_{1}\) and \(\sigma^2_{2}\) are often called the partial sill
and nugget, respectively.

@@

Other approaches, such as k-nearest-neighbors (Fix and Hodges, 1989; Ver
Hoef and Temesgen, 2013), random forests (Breiman, 2001), Bayesian
models (Chan-Golston et al., 2020), among others, could also be used to
obtain predictions for a mean or total from spatially correlated
responses of a finite population. Compared to the k-nearest-neighbors
and random forest approach, we prefer FBPK because it is model-based and
relies on theoretically-based variance estimators leveraging the model's
spatial covariance structure, whereas k-nearest-neighbors and random
forests use ad-hoc variance estimators (Ver Hoef and Temesgen, 2013).
Additionally, Ver Hoef and Temesgen (2013) studied compared FBPK,
k-nearest-neighbors, and random forests in a variety of spatial data
contexts, and FBPK tended to perform best. Compared to the Bayesian
approach, we prefer FPBK mostly because it is much more computationally
efficient.

@@

We used a simulation study to investigate performance of four
sampling-analysis combinations: IRS with a design-based analysis, called
``IRS-Design''; IRS with a model-based analysis, called ``IRS-Model'';
GRTS sampling with a design-based analysis, called ``GRTS-Design''; GRTS
sampling with a model-based analysis, called ``GRTS-Model''. These
combinations are also provided in Table \ref{tab:designanalysis}.

@@

Performance for the four sampling-analysis combinations was evaluated in
36 different simulation scenarios. The 36 scenarios resulted from the
crossing of three sample sizes, two location layouts, two response
types, and three proportions of dependent random error. The three sample
sizes (\(n\)) were \(n = 50, n = 100,\) and \(n = 200\). Samples were
always selected from a population size (\(N\)) of \(N = 900\). The two
location layouts (of the population units) were random and gridded.
Locations in the random layout were randomly generated inside the unit
square (\([0, 1] \times [0, 1]\)). Locations in the gridded layout were
placed on a fixed, equally spaced grid inside the unit square. The two
response types were normal and lognormal. For the normal response type,
the response was simulated using mean-zero random errors with the
exponential covariance (Equation\(~\)\ref{equation:expcov}) for varying
proportions of dependent random error. The proportion of dependent
random error is represented by
\(\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)\), where \(\sigma^2_1\) and
\(\sigma^2_2\) are the dependent random error variance (partial sill)
and independent random error variance (nugget), respectively, from
Equation\(~\)\ref{equation:expcov}. The total variance,
\(\sigma^2_1 + \sigma^2_2\), was always 2. The range was always
\(\sqrt{2} / 3\), which means that the correlation in the dependent
random error decayed to nearly zero at the largest possible distance
between two units in the domain. For the lognormal response type, the
response was first simulated using the same approach as for the normal
response type, except that the total variance was 0.6931 instead of 2.
The response was then exponentiated, yielding a random variable whose
total variance is 2. The lognormal responses were used to evaluate
performance of the sampling-analysis approaches for data that were
skewed (i.e., not normal).

\begin{table}[ht]
\centering
\begin{tabular}{r|lll}
   \hline
Sample Size (n) & 50 & 100 & 200 \\ 
  Location Layout & Random & Gridded & - \\ 
  Proportion of Dependent Error & 0 & 0.5 & 0.9 \\ 
  Response Type & Normal & Lognormal & - \\ 
   \hline
\end{tabular}
\caption{\label{tab:parmtab} Simulation scenario options. All combinations of sample size, location layout, response type, and proportion of dependent random error composed the 36 simulation scenarios. In each simualtion scenario, the total variance was two.} 
\end{table}

@@

The Environmental Protection Agency (EPA), states, and tribes
periodically conduct National Aquatic Research Surveys (NARS) in the
United States to assess the water quality of various bodies of water. We
will use data from the 2012 National Lakes Assessment (NLA), which
measures various aspects of lake health and water quality for lakes in
the contiguous United States (USEPA, 2012). Specifically, we will
analyze mercury concentration in lakes. Although we know the true mean
mercury concentration values for the 986 lakes from the 2012 NLA, we
will explore whether or not we obtain an adequately precise estimate for
the realized mean mercury concentration if we sample only 100 of the 986
lakes. For each of the four familiar sampling-analysis combinations
(IRS-Design, IRS-Model, GRTS-Design, and GRTS-Model), we estimate

@@

Fig. \ref{fig:figeff} shows the relative rMS(P)E of the four approaches
from Table \ref{tab:designanalysis} using the random location layout
with ``IRS-Design'' as the baseline.

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/rmspe_eff.jpeg}
  \caption{Relative rMS(P)E for the four sampling-analysis combinations. The rows indicate the proportion of dependent error and the columns indicate the response type.}
  \label{fig:figeff}
\end{figure}

The relative rMS(P)E is defined as \begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*} When there is no spatial correlation (Fig.
\ref{fig:figeff}, ``Prop DE: 0'' row), the four sampling-analysis
combinations have approximately equal rMS(P)E. So using the GRTS
sampling plan or a model-based analysis does not result in much, if any,
loss in efficiency compared to IRS-Design when there is no spatial
correlation. When there is spatial correlation (Fig. \ref{fig:figeff},
``Prop DE: 0.5'' and ``Prop DE: 0.9'' rows), GRTS-Model tends to perform
best, followed by GRTS-Design, IRS-Model, and finally IRS-Design, though
the difference in relative rMS(P)E among GRTS-Model, GRTS-Design, and
IRS-Model is relatively small. As the strength of spatial correlation
increases, the gap in rMS(P)E between IRS-Design and the other
sampling-analysis combinations widens. Finally we note that when there
is spatial correlation, IRS-Model outperforms IRS-Design by a large
margin, suggesting that the poor design properties of IRS are largely
mitigated by the model-based analysis. These conclusions are similar to
those observed in the grid location layout, so we omit a grid location
layout figure here. Tables for rMS(P)E in all 36 simulation scenarios
are provided in the supporting information.

We also studied 95\% interval coverage among the sampling-analysis
combinations. The design-based confidence intervals and model-based
prediction intervals were constructed using the normal distribution.
Justification for this comes from the asymptotic normality of means via
the Central Limit Theorem.

Fig. \ref{fig:figconf} shows the 95\% interval coverage for each of the
four sampling-analysis combinations in the random location layout.

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/coverage.jpeg}
  \caption{Interval coverage for the four sampling-analysis combinations. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid, black line in each plot represents 95\% coverage.}
  \label{fig:figconf}
\end{figure}

>>>>

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    The design-based and model-based approaches to frequentist statistical
    inference lie on fundamentally different foundations. In the
    design-based approach, inference depends on random sampling. In the
    model-based approach, inference depends on distributional assumptions.
    We compare the approaches for finite population spatial data.
  \item
    We provide relevant background for the design-based and model-based
    approaches and then study their performance using simulations and an
    analysis of real mercury concentration data. In the simulations, a
    variety of sample sizes, location layouts, dependence structures, and
    response types are considered. In the simualtions and real data
    analysis, the population mean is the parameter of interest and
    performance is measured using statistics like bias, squared error, and
    interval coverage.
  \item
    When studying the simulations and mercury concentration data, we found
    that regardless of the strength of spatial dependence in the data,
    sampling plans that incorporate spatial locations (spatially balanced
    samples) generally outperform sampling plans that ignore spatial
    locations (non-spatially balanced samples). We also found that
    model-based approaches tend to outperform design-based approaches,
    even when the data are skewed (and by consequence, the model-based
    distributional assumptions violated). The performance gap between the
    analysis approaches is small when spatially balanced samples are used
    but large when non-spatially balanced samples are used. This suggests
    that the sampling choice (whether to select a sample that is spatially
    balanced) is most important when performing design-based inference.
  \item
    There are many benefits and drawbacks to the design-based and
    model-based approaches for finite population spatial data that
    practitioners must consider when choosing between them. We provide
    relevant background contextualizing each approach and study their
    properties in a variety of scenarios, making recommendations for use
    based on the practitioner's goals.
  \end{enumerate}
  \end{abstract}

@@

There are two general approaches for using data to make frequentist
statistical inferences about a population: design-based and model-based.
When data cannot be collected for all units in a population (i.e.,
population units), data are collected on a subset of the population
units. This subset of population units is called a sample. In the
design-based approach, inferences about the underlying population are
informed via a probabilistic process that randomly assigns some
population units to be in the sample. Alternatively, in the model-based
approach, inferences are made from specific assumptions about the
underlying process generating the data. Each paradigm has a deep
historical context (Sterba, 2009) and its own set of benefits and
drawbacks (Hansen et al., 1983).

Though the design-based and model-based approaches apply to statistical
inference in a broad sense, we focus on comparing these approaches for
spatial data. We define spatial data as data that incorporates the
specific locations of the population units into either the design or
estimation process. De Gruijter and Ter Braak (1990) give an early
comparison of design-based and model-based approaches for spatial data,
quashing the belief that design-based approaches could not be used for
spatially correlated data. Since then, there have been several general
comparisons between design-based and model-based approaches for spatial
data (Brus and De Gruijter, 1997; Brus, 2021; Ver Hoef, 2002, 2008; Wang
et al., 2012). Cooper (2006) reviews the two approaches in an ecological
context before introducing a ``model-assisted'' variance estimator that
combines aspects from each approach. In addition to Cooper (2006), there
has been substantial research and development into estimators that use
both design and model-based principles (see e.g., Sterba (2009) and
Cicchitelli and Montanari (2012), and see Chan-Golston et al. (2020) for
a Bayesian approach).

Certainly comparisons between design-based and model-based approaches to
spatial data have been studied. But no numerical comparison has been
made between design-based approaches that incorporate spatial
information and model-based approaches. In this manuscript, we compare
design-based approaches that incorporate spatial information to
model-based approaches for finite population spatial data. A finite
population contains a finite number of population units (we assume the
finite number is known); an example is lakes (treated as a whole with
the lake centroid representing location) in the contiguous United
States. Though we manuscript focuses on finite populations, these
comparisons generalize to infinite populations as well. An infinite
population contains an infinite number of population units; an example
is locations within a single lake. In this manuscript we assume the
number of finite population units is known

The rest of the manuscript is organized as follows. In Section
\ref{sec:background}, we introduce and provide relevant background for
the design-based and model-based approaches to finite population spatial
data. In Section \ref{sec:mm}, we describe how we compare performance of
the approaches with a simulation study and an analysis of real data that
contains mercury concentration in lakes located in the contiguous United
States. In Section \ref{sec:results}, we present results from the
simulation study and the mercury concentration analysis. And in Section
\ref{sec:discussion}, we end with a discussion and provide directions
for future research.

@@

The design-based approach assumes the population is fixed. Randomness is
incorporated via the selection of units in a sampling frame. A sampling
frame is the set of all units available to be sampled. Units from the
sampling frame are selected as part of the sample according to a
sampling design, which assigns a positive probability of inclusion
(inclusion probability) to each unit from the sampling frame. These
inclusion probabilities are later used to analyze data. Some examples of
commonly used sampling designs include simple random sampling,
stratified random sampling, and cluster sampling.

When sampling designs incorporate spatial locations into sampling, we
call the resulting samples ``spatially balanced.'' One approach to
selecting spatially balanced samples is the Generalized Random
Tessellation Stratified (GRTS) algorithm (Stevens and Olsen, 2004),
which we discuss in more detail in Section \ref{subsec:spb_design}. When
sampling designs do not incorporate spatial locations into sampling, we
call the resulting samples ``non-spatially balanced.''

@@

The model-based approach assumes the data are a random realization of a
data-generating stochastic process. Randomness is incorporated through
distributional assumptions on this process. Strictly speaking,
randomness need not be incorporated through random sampling, though
Diggle et al. (2010) warn against preferential sampling. Preferential
sampling occurs when the process generating the data locations and the
process being modeled are not independent of one another. To guard
against preferential sampling, model-based approaches often still
implement some form of random sampling. When model-based approaches
implement random sampling, the inclusion probabilities are ignored when
analyzing the data (in contrast to the design-based approach, which
relies on these inclusion probabilities to analyze the data).

@@

We previously mentioned that the design-based approach can be used to
select spatially balanced samples (samples that incorporate spatial
locations of the population units). Spatially balanced samples are
useful because parameter estimates from these samples tend to vary less
than parameter estimates from samples that are not spatially balanced
(Barabesi and Franceschi, 2011; Benedetti et al., 2017; Grafström and
Lundström, 2013; Robertson et al., 2013; Stevens and Olsen, 2004; Wang
et al., 2013). The first spatially balanced sampling algorithm to see
widespread use was the Generalized Random Tessellation Stratified (GRTS)
algorithm (Stevens and Olsen, 2004). To quantify the spatial balance of
a sample, Stevens and Olsen (2004) proposed loss metrics based on
Voronoi polygons (Dirichlet Tessellations). After the GRTS algorithm was
developed, several other spatially balanced sampling algorithms emerged,
such as the Local Pivotal Method (Grafström et al., 2012; Grafström and
Matei, 2018), Spatially Correlated Poisson Sampling (Grafström, 2012),
Balanced Acceptance Sampling (Robertson et al., 2013),
Within-Sample-Distance Sampling (Benedetti and Piersimoni, 2017), and
Halton Iterative Partitioning Sampling (Robertson et al., 2018). In this
manuscript, we select spatially balanced samples using the Generalized
Random Tessellation Stratified (GRTS) algorithm because it has several
attractive properties. More specifically, the GRTS algorithm
accommodates finite and infinite sampling frames, equal, unequal, and
proportional (to size) inclusion probabilities, legacy (historical)
sampling (Foster et al., 2017), a minimum distance between units in a
sample, and replacement units (replacement units are population units
that can be sampled when a population unit originally selected can no
longer be sampled). The GRTS algorithm selects samples by utilizing a
particular mapping between two-dimensional and one-dimensional space
that preserves proximity relationships. Via this mapping, units in
two-dimensional space are partitioned using a hierarchical address. This
hierarchical address is used to map population units to a
one-dimensional line. On the one dimensional line, each population
unit's line length equals its inclusion probability. Then, a systematic
sample of population units is selected on the line and mapped back to
two-dimensional space, yielding the desired sample. Stevens and Olsen
(2004) provide more technical details.

After selecting a sample and collecting data, unbiased estimates of
population means and totals can be obtained using the Horvitz-Thompson
estimator (Horvitz and Thompson, 1952). If \(\tau\) is a population
total, the Horvitz-Thompson estimator for \(\tau\), denoted by
\(\hat{\tau}_{ht}\), is is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) is the value of the \(i\)th population unit in
the sample and \(\pi_i\) is the inclusion probability of the \(i\)th
population unit in the sample. An estimate of the population mean is
obtained by dividing \(\hat{\tau}_{ht}\) by \(N\), the number of
population units.

@@

Finite Population Block Kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
stochastic process. We summarize some of the basic principles of FBPK
next (for more technical details, see Ver Hoef (2008)). Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to use a sample to predict some linear function of the response
variable, \(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights (e.g, the
population mean is represented by a weights vector whose elements all
equal one). Denoting quantities that are part of the sampled population
units with a subscript \emph{s} and quantities that are part of the
unsampled population units with subscript \emph{u}, let

@@

FBPK assumes \(\bm{\delta}\) in Equation\(~\)\ref{equation:Zmarginal}
has mean-zero and a spatial dependence structure that can be modeled
using a covariance function. This covariance function is commonly
assumed to be non-negative, second-order stationary (depending only on
the distance between population units), isotropic (independent of
direction), and decay with distance between population units (Cressie,
1993). Henceforth, it is implied that we have made these same
assumptions regarding \(\bm{\delta}\), though Chiles and Delfiner
(1999), pp.~80-93 discuss covariance functions that are not second-order
stationary, not isotropic, or not either. A variety of flexible
covariance functions can be used to model \(\bm{\delta}\) (Cressie,
1993); one example is the exponential covariance function (for a
thorough list of spatial covariance functions, see Cressie (1993). The
\(i,j\)th element of the exponential covariance matrix,
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\), is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is the variance parameter quantifying
the variability that is dependent (coarse-scale), \(\sigma^2_{2}\) is
the variance parameter quantifying the variability that is independent
(fine-scale), \(\phi\) is the range parameter measuring the
distance-decay rate of the covariance, and \(h_{i,j}\) is the Euclidean
distance between population units \(i\) and \(j\). The proportion of
variability attributable to dependent random error is
\(\sigma^2_{1} / (\sigma^2_{1} + \sigma^2_{2})\). Similarly, the
proportion of variability attributable to independent random error is
\(\sigma^2_{2} / (\sigma^2_{1} + \sigma^2_{2})\). Finally we note that
\(\sigma^2_{1}\) and \(\sigma^2_{2}\) are often called the partial sill
and nugget, respectively.

@@

Other approaches, such as k-nearest-neighbors (Fix and Hodges, 1989; Ver
Hoef and Temesgen, 2013) and random forests (Breiman, 2001), among
others, could also be used to obtain predictions for a mean or total
from spatially correlated responses of a finite population. Compared to
the k-nearest-neighbors and random forest approach, we prefer FBPK
because it is model-based and relies on theoretically-based variance
estimators leveraging the model's spatial covariance structure, whereas
k-nearest-neighbors and random forests use ad-hoc variance estimators
(Ver Hoef and Temesgen, 2013). Additionally, Ver Hoef and Temesgen
(2013) studied compared FBPK, k-nearest-neighbors, and random forests in
a variety of spatial data contexts, and FBPK tended to perform best.

@@

We used a simulation study to investigate performance of four
sampling-analysis combinations: IRS sampling with a design-based
analysis, called ``IRS-Design''; IRS sampling with a model-based
analysis, called ``IRS-Model''; GRTS sampling with a design-based
analysis, called ``GRTS-Design''; GRTS sampling with a model-based
analysis, called ``GRTS-Model''. These combinations are also provided in
Table \ref{tab:designanalysis}.

@@

Performance for the four sampling-analysis combinations was evaluated in
36 different simulation scenarios. The 36 scenarios resulted from the
crossing of three sample sizes, two location layouts, two response
types, and three proportions of dependent random error. The three sample
sizes (\(n\)) were \(n = 50, n = 100,\) and \(n = 200\). Samples were
always selected from a population size (\(N\)) of \(N = 900\). The two
location layouts (of the population units) were random and gridded.
Locations in the random layout were randomly generated inside the unit
square (\([0, 1] \times [0, 1]\)). Locations in the gridded layout were
placed on a fixed, equally spaced grid inside the unit square. The two
response types were normal and lognormal. For the normal response type,
the response was simulated using mean-zero random errors with the
exponential covariance (Equation\(~\)\ref{equation:expcov}) for varying
proportions of dependent random error. The proportion of dependent
random error is represented by
\(\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)\), where \(\sigma^2_1\) and
\(\sigma^2_2\) are the dependent random error variance (partial sill)
and independent random error variance (nugget), respectively, from
Equation\(~\)\ref{equation:expcov}. The total variance,
\(\sigma^2_1 + \sigma^2_2\), was always 2. The range was always
\(\sqrt{2} / 3\), which means that the correlation in the dependent
random error decayed to nearly zero at the largest possible distance
between two population units in the domain. For the lognormal response
type, the response was first simulated using the same approach as for
the normal response type, except that the total variance was 0.6931
instead of 2. The response was then exponentiated, yielding a lognormal
random variable whose total variance is 2. The lognormal responses were
used to evaluate performance of the sampling-analysis approaches for
data that were skewed (i.e., not normal).

\begin{table}[ht]
\centering
\begin{tabular}{r|lll}
   \hline
Sample Size (n) & 50 & 100 & 200 \\ 
  Location Layout & Random & Gridded & - \\ 
  Proportion of Dependent Error & 0 & 0.5 & 0.9 \\ 
  Response Type & Normal & Lognormal & - \\ 
   \hline
\end{tabular}
\caption{\label{tab:parmtab} Simulation scenario options. All combinations of sample size, location layout, response type, and proportion of dependent random error composed the 36 simulation scenarios. In each simualtion scenario, the total variance was 2.} 
\end{table}

@@

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the contiguous
United States. We will use data from the 2012 National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality
(USEPA, 2012). Specifically, we will analyze mercury concentration in
lakes. Although we can calculate the true mean mercury concentration
values for the 986 lakes from the 2012 NLA, we will explore whether or
not we obtain an adequately precise estimate for the realized mean
mercury concentration if we sample only 100 of the 986 lakes. For each
of the four familiar sampling-analysis combinations (IRS-Design,
IRS-Model, GRTS-Design, and GRTS-Model), we estimate (design-based) or
predict (model-based) the mean mercury concentration and construct 95\%
confidence (design-based) or prediction (model-based) intervals from
this sample of 100 lakes, which we compare to the actual mean from all
986 lakes.

@@

Fig. \ref{fig:figeff} shows the relative rMS(P)E of the four approaches
from Table \ref{tab:designanalysis} using the random location layout
with ``IRS-Design'' as the baseline. The relative rMS(P)E is defined as
\begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*} When there is no spatial correlation (Fig.
\ref{fig:figeff}, ``Prop DE: 0'' row), the four sampling-analysis
combinations have approximately equal rMS(P)E. So using the GRTS
sampling plan or a model-based analysis does not result in much, if any,
loss in efficiency compared to IRS-Design when there is no spatial
correlation. When there is spatial correlation (Fig. \ref{fig:figeff},
``Prop DE: 0.5'' and ``Prop DE: 0.9'' rows), GRTS-Model tends to perform
best, followed by GRTS-Design, IRS-Model, and finally IRS-Design, though
the difference in relative rMS(P)E among GRTS-Model, GRTS-Design, and
IRS-Model is relatively small. As the strength of spatial correlation
increases, the gap in rMS(P)E between IRS-Design and the other
sampling-analysis combinations widens. Finally we note that when there
is spatial correlation, IRS-Model outperforms IRS-Design by a large
margin, suggesting that the poor design properties of IRS are largely
mitigated by the model-based analysis. These conclusions are similar to
those observed in the grid location layout, so we omit a grid location
layout figure here. Tables for rMS(P)E in all 36 simulation scenarios
are provided in the supporting information.

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/rmspe_eff.jpeg}
  \caption{Relative rMS(P)E in the simulation study for the four sampling-analysis combinations. The rows indicate the proportion of dependent error and the columns indicate the response type.}
  \label{fig:figeff}
\end{figure}
